---
title: "Getting started with odrlITR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with odrlITR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Method and coding convention

`odrlITR` internally represents treatment by \(A\in\{-1,+1\}\), with larger
outcomes preferred. It estimates

\[
e(x)=E(A\mid X=x), \qquad m(x)=E(Y\mid X=x),
\]

and forms the pooled out-of-fold score

\[
\widehat Z_i=\{A_i-\widehat e(X_i)\}
              \{Y_i-\widehat m(X_i)\}.
\]

The direct learners minimize weighted policy error, equivalently maximizing
\(\sum_i \widehat Z_i d(X_i)\). The surrogate learners minimize a weighted
hinge, exponential, logistic, squared-hinge, or user-specified differentiable
margin loss using labels `sign(Z)` and weights `abs(Z)`. For signed margin
\(t\), `loss = "exponential"` uses \(\phi(t)=\exp(-t)\), with a convex linear
continuation only below the extreme margin \(t=-30\) to avoid overflow.

For conditioning and comparable penalty grids, the second stage divides
\(\widehat Z\) by its sample mean absolute value. This is a positive rescaling:
it leaves the unpenalized binary policy objective unchanged. Reported training
criteria are evaluated on the original score scale.

## A reproducible example

```{r simulate}
library(odrlITR)
set.seed(1)
dat <- odrl_simulate(n = 300, boundary = "tree", seed = 1)
known <- odrl_nuisance_user(
  m = dat$m,
  pi = dat$pi,
  source = "known simulation truth",
  out_of_fold = TRUE
)
```

The following tree fit is evaluated only when the optional engine is present.

```{r tree}
if (requireNamespace("policytree", quietly = TRUE)) {
  tree_fit <- odrl(dat$x, dat$a, dat$y, learner = "tree",
                   nuisance = known,
                   control = odrl_control(tree_depth = 2, seed = 10))
  tree_fit
  head(predict(tree_fit, dat$x))
}
```

The package also includes a self-contained bounded-hinge Gaussian-kernel
learner:

```{r svm}
svm_fit <- odrl(
  dat$x, dat$a, dat$y,
  learner = "svm", loss = "hinge", nuisance = known,
  control = odrl_control(
    svm_rbf_multiplier = c(0.5, 1, 2),
    svm_radius = c(0.5, 1),
    svm_folds = 3,
    seed = 11
  )
)
svm_fit$policy$selected
range(predict(svm_fit, dat$x, type = "score"))
```

Set `svm_kernel` to `"linear"`, `"rbf"`, or `"polynomial"`; a custom
function or a list containing `name`, `fun`, and `args` defines another kernel.
The globally bounded RKHS-radius construction is specific to Gaussian/RBF
hinge fitting. Use `svm_hinge_mode = "regularized"` for ordinary hinge loss
with any supported kernel. For example:

```{r svm-flexible, eval=FALSE}
polynomial_fit <- odrl(
  x, a, y,
  learner = "svm", loss = "squared_hinge",
  control = odrl_control(
    svm_kernel = "polynomial",
    svm_polynomial_degree = c(2, 3),
    svm_penalty = c(0.01, 0.1, 1),
    svm_hinge_mode = "regularized"
  )
)

smooth_loss <- list(
  name = "softplus_margin",
  value = function(margin) {
    z <- 1 - margin
    pmax(z, 0) + log1p(exp(-abs(z)))
  },
  gradient = function(margin) -plogis(1 - margin)
)
custom_fit <- odrl(x, a, y, learner = "svm", loss = smooth_loss)
```

Custom kernels must be symmetric and positive semidefinite. Custom losses must
return finite nonnegative values and finite derivatives; the caller remains
responsible for convexity when a convex RKHS problem is intended.

The Gaussian learner constructs and stores a dense \(n \times n\) kernel
matrix, so both fitting time and memory scale at least quadratically with the
training sample size. Prediction also evaluates the kernel against every
training observation. For large data sets, first benchmark this learner on a
representative subset or use the affine, tree, or ReLU policy class.

For regularized kernel-surrogate fits, both the constant offset and the RKHS
component are included in the quadratic penalty.

## Cross-fitted Super Learner

With `nuisance = NULL` (the default), `odrl()` calls
`odrl_nuisance_sl()`. Super Learner's internal CV is nested inside the outer
cross-fitting loop. The number of outer folds is configurable. By default,
the package stratifies them by treatment while balancing total fold sizes to
differ by at most one row. For example:

```{r superlearner, eval=FALSE}
fit <- odrl(
  x, treatment, outcome,
  learner = "relu", loss = "logistic",
  nuisance_folds = 5,
  sl_inner_folds = 5,
  sl.library = c("SL.mean", "SL.glm", "SL.glmnet")
)
```

To use prespecified outer folds, supply the same row-aligned identifiers to
`nuisance_fold_id`; their labels are normalized and retained for auditing.
Each outer training set must contain at least two observations and, when the
propensity is estimated, at least two observations from each treatment arm:

```{r custom-folds, eval=FALSE}
fit <- odrl(
  x, treatment, outcome,
  learner = "tree",
  nuisance_fold_id = study_fold,
  sl.library = c("SL.mean", "SL.glm", "SL.glmnet")
)
```

The two nuisance libraries may be specified separately through
`sl.library.pi` and `sl.library.m`. For a randomized trial, `known_pi` (or
`known_e`) skips propensity fitting but preserves outer cross-fitting for the
marginal outcome regression:

```{r known-propensity, eval=FALSE}
fit <- odrl(
  x, treatment, outcome,
  learner = "tree",
  known_pi = 0.5,
  sl.library.m = c("SL.mean", "SL.glm", "SL.glmnet")
)
```

Custom wrappers can be supplied in `sl.library`, with `sl_env` identifying the
environment in which Super Learner should resolve them.

The returned nuisance object preserves the libraries, coefficients, actual
inner-fold counts, and captured warning messages for each outer fold in
`nuisance$fits`; `nuisance$warnings` is the deduplicated warning ledger.

## Cross-fitted parametric nuisances

For a built-in parametric first stage, set `nuisance = "parametric"` (or
`"glm"`). Each outer training fold fits a main-effects binomial logistic model
for \(P(A=+1\mid X)\) and a Gaussian linear model for \(E(Y\mid X)\), then
predicts its holdout rows:

```{r parametric, eval=FALSE}
fit <- odrl(
  x, treatment, outcome,
  learner = "linear",
  nuisance = "parametric",
  nuisance_folds = 4
)
```

Supply interactions or transformations as columns of `x` when they belong to
the intended parametric basis. `known_pi` and `known_e` work with this engine
as well, in which case only the marginal outcome model is cross-fitted.

## User nuisance contract

Externally estimated predictions must align with training rows. Unless known
by design, they should be out of fold. The package warns when an object is not
marked out of fold:

```{r user, eval=FALSE}
nuisance <- odrl_nuisance_user(
  m = m_hat_oof,
  e = e_hat_oof,
  fold_id = fold_id,
  source = "external learner",
  out_of_fold = TRUE
)
fit <- odrl(x, a, y, learner = "linear", nuisance = nuisance)
```

No clipping is applied unless `propensity_bounds` is set explicitly. The
fitted nuisance object records the raw probability range and clipped fraction.

## Optimization diagnostics

The direct affine learner may stop with a feasible incumbent before proving
the requested MIP gap. The fitted object records the distinction:

```{r linear, eval=FALSE}
fit <- odrl(
  x, a, y, learner = "linear", nuisance = nuisance,
  control = odrl_control(
    linear_time_limit = 600,
    linear_relative_gap = 0.01,
    linear_absolute_gap = 0.001,
    linear_node_limit = 100000,
    linear_threads = 1,
    linear_require_gap = FALSE
  )
)
fit$policy$diagnostics
```

`linear_require_gap = TRUE` requires every requested relative and absolute gap
criterion. Node and objective limits, thread count, logging, and additional
nonconflicting HiGHS options are also configurable. To request a numerical
global-optimum certificate within the bounded-margin standardized affine
class, combine it with `linear_relative_gap = 0` and inspect
`certified_global_optimum` in the returned diagnostics.

`coef(fit)` returns affine coefficients on the original encoded-covariate
scale; use `coef(fit, standardized = TRUE)` for the internally standardized
coefficients. For trees, `split.step = 1` searches all candidate split points
admitted by `policytree`; larger values deliberately thin that candidate set.
`tree_backend` may instead be a list containing named `fit` and `predict`
callbacks, and `tree_options` is passed to the selected engine.

## Neural policy classes

The native neural engine supports an affine candidate, arbitrary hidden-layer
width vectors, and ReLU, leaky-ReLU, tanh, sigmoid, or linear activations:

```{r neural, eval=FALSE}
fit <- odrl(
  x, a, y,
  learner = "relu", loss = "logistic",
  control = odrl_control(
    relu_architectures = list(integer(), 8L, c(16L, 8L)),
    relu_activation = c("relu", "tanh"),
    relu_decay = c(0.001, 0.01, 0.1),
    relu_selection = "one_se"
  )
)
```

`relu_backend` can instead be a list containing `name`, `fit`, and `predict`
callbacks, allowing another neural-network package to perform the numerical
optimization while `odrlITR` retains score construction, tuning, and audit
metadata. A custom loss can be supplied as the same `value`/`gradient` list
used by the SVM engine or as a function returning `list(loss, gradient)`.

Hard-tanh bounding gives the native hinge score the universal-orthogonality
property, but this implementation does not claim the manuscript's sparse
deep-network entropy rate.
