---
title: "Getting started with odrl"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with odrl}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Method and coding convention

`odrl` internally represents treatment by (A\in\{-1,+1\}), with larger
outcomes preferred. It estimates

\[
e(x)=E(A\mid X=x), \qquad m(x)=E(Y\mid X=x),
\]

and forms the pooled out-of-fold score

\[
\widehat Z_i=\{A_i-\widehat e(X_i)\}
              \{Y_i-\widehat m(X_i)\}.
\]

The direct learners minimize weighted policy error, equivalently maximizing
\(\sum_i \widehat Z_i d(X_i)\). The surrogate learners minimize weighted hinge
or logistic loss using labels `sign(Z)` and weights `abs(Z)`.

For conditioning and comparable penalty grids, the second stage divides
\(\widehat Z\) by its sample mean absolute value. This is a positive rescaling:
it leaves the unpenalized binary policy objective unchanged. Reported training
criteria are evaluated on the original score scale.

## A reproducible example

```{r simulate}
library(odrl)
set.seed(1)
dat <- odrl_simulate(n = 300, boundary = "tree", seed = 1)
known <- odrl_nuisance_user(
  m = dat$m,
  pi = dat$pi,
  source = "known simulation truth",
  out_of_fold = TRUE
)
```

The following tree fit is evaluated only when the optional engine is present.

```{r tree}
if (requireNamespace("policytree", quietly = TRUE)) {
  tree_fit <- odrl(dat$x, dat$a, dat$y, learner = "tree",
                   nuisance = known,
                   control = odrl_control(tree_depth = 2, seed = 10))
  tree_fit
  head(predict(tree_fit, dat$x))
}
```

The package also includes a self-contained bounded-hinge Gaussian-kernel
learner:

```{r svm}
svm_fit <- odrl(
  dat$x, dat$a, dat$y,
  learner = "svm", loss = "hinge", nuisance = known,
  control = odrl_control(
    svm_rbf_multiplier = c(0.5, 1, 2),
    svm_radius = c(0.5, 1),
    svm_folds = 3,
    seed = 11
  )
)
svm_fit$policy$selected
range(predict(svm_fit, dat$x, type = "score"))
```

The Gaussian learner constructs and stores a dense \(n \times n\) kernel
matrix, so both fitting time and memory scale at least quadratically with the
training sample size. Prediction also evaluates the kernel against every
training observation. For large data sets, first benchmark this learner on a
representative subset or use the affine, tree, or ReLU policy class.

For logistic kernel fits, both the constant offset and the RKHS component are
included in the quadratic penalty.

## Cross-fitted Super Learner

With `nuisance = NULL` (the default), `odrl()` calls
`odrl_nuisance_sl()`. Super Learner's internal CV is nested inside the outer
cross-fitting loop. For example:

```{r superlearner, eval=FALSE}
fit <- odrl(
  x, treatment, outcome,
  learner = "relu", loss = "logistic",
  nuisance_folds = 5,
  sl_inner_folds = 5,
  sl.library = c("SL.mean", "SL.glm", "SL.glmnet")
)
```

The two nuisance libraries may be specified separately through
`sl.library.pi` and `sl.library.m`. For a randomized trial, `known_pi` (or
`known_e`) skips propensity fitting but preserves outer cross-fitting for the
marginal outcome regression:

```{r known-propensity, eval=FALSE}
fit <- odrl(
  x, treatment, outcome,
  learner = "tree",
  known_pi = 0.5,
  sl.library.m = c("SL.mean", "SL.glm", "SL.glmnet")
)
```

Custom wrappers can be supplied in `sl.library`, with `sl_env` identifying the
environment in which Super Learner should resolve them.

The returned nuisance object preserves the libraries, coefficients, actual
inner-fold counts, and captured warning messages for each outer fold in
`nuisance$fits`; `nuisance$warnings` is the deduplicated warning ledger.

## User nuisance contract

Externally estimated predictions must align with training rows. Unless known
by design, they should be out of fold. The package warns when an object is not
marked out of fold:

```{r user, eval=FALSE}
nuisance <- odrl_nuisance_user(
  m = m_hat_oof,
  e = e_hat_oof,
  fold_id = fold_id,
  source = "external learner",
  out_of_fold = TRUE
)
fit <- odrl(x, a, y, learner = "linear", nuisance = nuisance)
```

No clipping is applied unless `propensity_bounds` is set explicitly. The
fitted nuisance object records the raw probability range and clipped fraction.

## Optimization diagnostics

The direct affine learner may stop with a feasible incumbent before proving
the requested MIP gap. The fitted object records the distinction:

```{r linear, eval=FALSE}
fit <- odrl(
  x, a, y, learner = "linear", nuisance = nuisance,
  control = odrl_control(
    linear_time_limit = 600,
    linear_relative_gap = 0.01,
    linear_require_gap = FALSE
  )
)
fit$policy$diagnostics
```

`linear_require_gap = TRUE` requires the requested relative gap. To request a
numerical global-optimum certificate within the bounded-margin standardized
affine class, combine it with `linear_relative_gap = 0` and inspect
`certified_global_optimum` in the returned diagnostics.

`coef(fit)` returns affine coefficients on the original encoded-covariate
scale; use `coef(fit, standardized = TRUE)` for the internally standardized
coefficients. For trees, `split.step = 1` searches all candidate split points
admitted by `policytree`; larger values deliberately thin that candidate set.

The ReLU engine implemented here is a regularized one-hidden-layer subclass
with an optional affine candidate. Hard-tanh bounding gives its hinge score the
universal-orthogonality property, but this implementation does not claim the
manuscript's sparse deep-network entropy rate.
