---
title: "Getting started with odrlITR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with odrlITR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Method

`odrlITR` represents treatment by (A\in\{-1,+1\}) and estimates

\[
e(x)=E(A\mid X=x), \qquad m(x)=E(Y\mid X=x).
\]

Cross-fitted nuisance predictions define the second-stage score

\[
\widehat Z_i=\{A_i-\widehat e(X_i)\}
              \{Y_i-\widehat m(X_i)\}.
\]

Direct learners maximize the empirical policy criterion
\(\sum_i\widehat Z_i d(X_i)\). Surrogate learners use `sign(Z)` as the label
and `abs(Z)` as the observation weight. The package divides the score by its
sample mean absolute value while fitting the policy; reported policy criteria
remain on the original scale.

## A complete example

The built-in parametric nuisance and kernel SVM paths require no optional
packages.

```{r fit}
library(odrlITR)

dat <- odrl_simulate(n = 200, boundary = "linear", seed = 1)

fit <- odrl(
  dat$x, dat$a, dat$y,
  learner = "svm",
  loss = "hinge",
  nuisance = "parametric",
  nuisance_folds = 5,
  control = odrl_control(
    svm_penalty = c(0.05, 0.2),
    svm_rbf_multiplier = c(0.5, 1),
    svm_folds = 2,
    seed = 2
  )
)

head(predict(fit, dat$x))
range(predict(fit, dat$x, type = "score"))
fit$policy$selected
```

`predict(fit)` returns treatment in the coding used for fitting.
`predict(fit, type = "score")` returns the decision score. For factor or
character treatment, use `positive` to identify the `+1` level.

## Bounded hinge SVM

Example 3 of the paper uses two steps:

1. fit a penalized RKHS score by minimizing the empirical double-residual
   hinge loss;
2. clip the fitted score with
   \(T_1(u)=\max\{-1,\min(1,u)\}\).

Clipping preserves the sign of the score and cannot increase hinge loss. The
result is bounded by one, although it need not itself be an RKHS function. The
paper uses this bounded score for its universal Neyman-orthogonality result.

This is the default for hinge loss with the Gaussian/RBF kernel. Use explicit
bounded mode for another supported kernel:

```{r linear-bounded, eval=FALSE}
linear_fit <- odrl(
  dat$x, dat$a, dat$y,
  learner = "svm", loss = "hinge",
  nuisance = "parametric",
  control = odrl_control(
    svm_kernel = "linear",
    svm_hinge_mode = "bounded",
    svm_penalty = c(0.01, 0.1, 1)
  )
)
```

Use `svm_hinge_mode = "regularized"` for an unclipped hinge score. The SVM
penalizes the kernel component and a constant offset, which is equivalent to
using the augmented kernel (K+1). The fitted object records the selected
penalty, score norm, convergence code, and primal-dual gap.

## Nuisance choices

`nuisance = "parametric"` cross-fits a logistic propensity model and a linear
marginal-outcome model. Include transformations or interactions as columns of
`x` when they are part of the intended models.

The default, `nuisance = NULL`, uses cross-fitted Super Learner:

```{r superlearner, eval=FALSE}
fit <- odrl(
  dat$x, dat$a, dat$y,
  learner = "svm",
  nuisance_folds = 5,
  sl_inner_folds = 5,
  sl.library = c("SL.mean", "SL.glm", "SL.glmnet")
)
```

Use `sl.library.pi` and `sl.library.m` to set separate libraries. Supply
`known_pi` or `known_e` in a randomized trial. `nuisance_fold_id` accepts a
row-aligned outer-fold assignment.

User-supplied nuisance predictions are accepted through
`odrl_nuisance_user()`:

```{r user-nuisance, eval=FALSE}
nuisance <- odrl_nuisance_user(
  m = m_hat_oof,
  e = e_hat_oof,
  fold_id = fold_id,
  source = "external learner",
  out_of_fold = TRUE
)

fit <- odrl(dat$x, dat$a, dat$y, learner = "svm", nuisance = nuisance)
```

Propensity clipping is off by default because the ODRL score does not divide
by the propensity. Set `propensity_bounds` when clipping is desired.

## Policy learners

| `learner` | fit | optional package |
|:--|:--|:--|
| `"tree"` | exact shallow policy tree | `policytree` |
| `"linear"` | exact affine rule by mixed-integer programming | `highs` |
| `"svm"` | kernel or finite-series surrogate | none |
| `"relu"` | native neural surrogate or external backend | `nnet` for that backend |

Tree depth is a maximum: zero fits a constant rule, one a stump, and two
allows a root split followed by child splits. Linear-solver status and gap
information are stored in `fit$policy$diagnostics`.

SVM and neural learners support hinge, exponential, logistic, squared hinge,
and custom differentiable signed-margin losses. Kernel choices include
`"linear"`, `"rbf"`, and `"polynomial"`; a function or named specification
can define a custom kernel.

```{r other-learners, eval=FALSE}
tree_fit <- odrl(
  dat$x, dat$a, dat$y, learner = "tree", nuisance = "parametric",
  control = odrl_control(tree_depth = 2)
)

neural_fit <- odrl(
  dat$x, dat$a, dat$y,
  learner = "relu", loss = "logistic", nuisance = "parametric",
  control = odrl_control(relu_preset = "fast")
)
```

## Finite-series scores

`odrl_series_kernel()` defines explicit Legendre, Fourier, B-spline,
Haar-wavelet, and local-polynomial features. The feature map is learned within
each policy-training fold and reused for validation and prediction.

```{r series, eval=FALSE}
series <- odrl_series_kernel(
  basis = "legendre",
  legendre_degree = c(1, 2, 3),
  combine = "anova",
  interaction_order = 2,
  max_features = 1000
)

series_fit <- odrl(
  dat$x, dat$a, dat$y,
  learner = "svm", loss = "logistic", nuisance = "parametric",
  control = odrl_control(
    svm_kernel = series,
    svm_penalty = c(0.01, 0.1, 1)
  )
)
```

The additive basis omits interactions. `combine = "anova"` adds interactions
up to `interaction_order`; `"total_degree"` builds a total-degree Legendre
basis; and `"tensor"` requests a full tensor product. Gaussian and custom
kernels retain a dense training kernel matrix, while finite-series learners
fit their explicit features directly.

## Inspecting a fit

The main components are:

- `fit$nuisance`: cross-fitted first-stage predictions and metadata;
- `fit$score`: raw and normalized double-residual scores;
- `fit$policy`: selected tuning values and learner diagnostics;
- `fit$training`: fitted actions and empirical policy criterion.

Use `summary(fit)`, `fitted(fit)`, and `coef(fit)` where applicable. The
function reference contains the full set of learner and backend controls.
